---
title: "Quiz2Review"
author: "Xuanpei Ouyang"
date: "October 13, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Slide 1
```{r}
# loead the package
library('ggplot2')  

# search a specific function
help("+")
?"+"
help("vector")
?vector

# returns a list of functions and packages related with the searched term
??matrix
help.search("matrix")

# returns a list of functions containing the searched term
apropos("mean")

# return all the functions in a given package
help(package = "ggplot2")

# get a list of packages contained in R
# by specifying the lib.loc argument
library(lib.loc = .Library)

# find which package a function belongs
find("mean")
find("ggplot")
```

## Slide 2

```{r}
# function c() - concatenate
x = c(1,2,3,4,5)
```

#### Atomic structure: vector, matrix, array
#### Non-atomic structure: list, data.frame

```{r}
# vector - atomic structure - coerce
c(1,2,3,"four","five") # coerce numbers into string
c(TRUE, FALSE, 3, 4) # coerce logical into numbers
c(TRUE, FALSE, "TRUE", "FALSE") #coerce logical into string
c(1L, -0.5) # coerce integer to real number
c(1L, -0.5, 3+5i) # coerce real number to complex

# vector with named elements
x = c(a = 1, b = 2, c = 3)
length(x)
mode(x)

# convert vectors to arrays
x = 1:8
class(x) # 1 dim is class integer
dim(x) = c(2,4) # 2 dimension
x
class(x) # 2 dim is class matrix
dim(x) = c(2,2,2) # 3 dimension
x
class(x) # 3 dim is class array

# Matrix
matrix(x, nrow = 2, ncol = 4, byrow = TRUE)

hp = c("A","B","C","D","E","F")
matrix(hp, nrow = 3, byrow =TRUE)

# Array
ar = array(c(1:8), dim = c(2,2,2))
ar

# Factor - used for categorical data
# Mode - numeric (internally stored as integers)
factor(c("A","B","C")) # different values are levels
```

#### Lists  
Group data into a one-dimensional set  
Can store all kinds of objects  
```{r}
list(1,2,3)
# list can contain any type of data object (even list)
list(
  c("A","B","C"),
  matrix(1:6, nrow = 2, ncol = 3),
  factor(c("a","b","c")))
# elements in a list can be named
list(
  row1 = c(1,2,3),
  row2 = c(4,5,6))

list("Xuanpei",NULL,"Ouyang")
```

#### Data Frame  
The primary data structure for tabular datasets  
Two-dimensional versino of a list  
A data frame is stored as a list  
Give data.frame() any number of vectors, each seperated with a comma and it will turn each vector into a column of the new data frame  
All vectors should be of the same length  
By default, data.frame() converts strings into factors
```{r}
df1 = data.frame(
  name = c("hydrogen","oxygen"),
  number = c(1,7))
df1

df2 = data.frame(
  state = c("California","New York", "Texas"),
  abbreviation = c("CA","NY","TX"),
  capital = c("Sacramento", "Albany","Austin"),
  area = c(163707, 54475, 268601),
  stringsAsFactors = FALSE)
df2
```
#### Selecting Values
Use brackets [ ] to extract values  
Use as many indices as dimensions in the object  
Each index is seperated by comma  
Index can be numbers, logicals, or names
```{r}
vec = 1:5
names(vec) = letters[1:5]

# get the first element in vec
vec[1]
vec["a"]

# get first three elements in vec
vec[1:3]
vec[c(1,2,3)]
vec[c("a","b","c")]

# get all values except the first one
vec[-1]

# get all values except the 2nd and 4th
vec[-c(2,4)]

# get the first element
vec[c(TRUE, FALSE, FALSE, FALSE, FALSE)]
vec[!c(FALSE, TRUE, TRUE, TRUE, TRUE)]

# get the first element 3 times
vec[c(1,1,1)]
vec[c("a","a","a")]
```

#### Extracting values of two-dimensinal objects
```{r}
# value in row 1 and column 1
df2[1,1]

# all the values in row 1
df2[1,]

# all the values in row 1 to 2
df2[1:2,]

# all the values in column 1
df2[,1]

# all the values in column 2 to 4
df2[,2:4]

# extracting everything except 3rd row
df2[-3,]

# extracting everything except 1st and 3rd row
df2[-c(1,3),]

# extracting everying except 2nd column
df2[,-2]

# extracting everythinge except 1st and 2nd columns
df2[,-c(1,2)]

# extract column 1 twice
df2[,c(1,1)]

# extracting columns by name
df2[,"state"]

# When select one column from a two-dimensional array, R will return a vector.
# To get a column output, use the argument drop = FALSE
df2[ , 1, drop = FALSE]
```

#### Dollar signs
Selecting a column of a data frame using its name  
Work with data frame and a named list  
If doesn't exist, will return NULL
```{r} 
df2$state
df2$"state"
df2$`state`

df2$numbers
```
#### Double brackets  
Get access to the individual elements - double brackets followed by single brackets  
Work with data frame and list
```{r}
# get the 2nd column
df2[[2]]

# get the third element in 2nd column
df2[[2]][3]
```
| object | notation | example |
|:-------: | :-------: |:-------: |
| vector | [ ] | v[2] |
| factor | [ ] | g[2] |
| matrix | [ , ] | m[1,3]  |
| array | [ , , ] | ar[1,2,3] |
| array | [ , , , ] | ar[1,2,3,4] |
| list | [ ] | list[3] |
| list | [[ ]] | list[[3]] |
| list | $ | list$name |
| data frame | [ , ] | df[1, 2] | 
| data frame | $ | df$name |

## Slide 3
```{r}
# initialize vectors of a particular class 
integer(5) # 0
numeric(5) # 0
complex(5) # 0+0i
logical(5) # FALSE
character(5) # ""

# check if input is a integer
is.integer(3L)

# convert the input to a integer
as.integer(FALSE) # 0
as.integer("2") # 2
as.integer("f") # NA
```

#### Numeric Vector of Sequence
```{r}
1:5 # 1,2,3,4,5
-7:-4 # -7, -6, -5, -4
1:-1 # 1, 0, -1

# seq()
seq(1)
seq(from = 1, to = 5)
seq(from = 3, to = -3)
seq(from = 1, to = 10, by = 2)
seq(from = 1, to = 10, length.out = 5)

# seq_along() 
# returns a sequence of integers of the same length as its argument
seq_along(c(4,2,3)) # 1,2,3

# seq_len() 
# generates a sequence from 1 to the value provided
seq_len(5)

# rep()
# create vector with replicating numbers
rep(1, times = 5) # 1,1,1,1,1
rep(1:3, times = 2) # 1,2,3,1,2,3
rep(1:3, each = 2) # 1,1,2,2,3,3
rep(1:3, length.out = 5) # 1,2,3,1,2

# generate random vectors
runif(5) # random numbers from uniform distribution
rnorm(5) # random numbers from normal distribution
runif(n = 5, min = 0, max = 1)
rnorm(n = 5, mean = 1, sd = 1)

# sampled vectors
sample(1:10, size = 10) # without replacememt
sample(1:10, size = 10, replace = TRUE) # with replacement
```

#### Vector Functions
```{r}
# return the length of a vector
length(1:10)

# sorting elements
sort(10:1) # by default increasing order
sort(10:1, decreasing = TRUE)  

# reverse elements in a vector
rev(1:10)
  
# return the position of sorted elements
order(c(1,4,2,3))
  
# return the unique elements
unique(c(1,2,3,3,4,4))
 
# return logical vector, true indicating that this number appears before
duplicated(c(1,2,3,3,4,4))
```

#### Arithmetic Operators
```{r}
x = 2
y = 3
+ x # positive
- x # negative
x ^ y # power
x %% y # module
x %/% y # integer division
```

#### Math Functions
```{r}
abs(c(-1,1))
sign(c(-2,2))
round(3.14159)
log10(10)
```

#### Vectorized Operations
```{r}
c(1,1,1) * c(2,2,2)
log(c(1,2,3))
sqrt(1:3)
```

#### Recycling
R will give warning if the longer object length is not a multiple of
shorter object length  
The recycling rule states that the shorter vector is replicated enough times  
so that the result has the length of the longer vector
```{r}
c(1,2) + c(1,2,3) # will be 2,4,4
c(1,2) + c(1,2,3,4) # will be 2,4,4,6
```

#### Comparison operator
When comparing vectors of different type, one is coerced to the type of the  
other, the decreasing order of precedence is  
character > complex > numeric > integer > logical
```{r}
'5' == 5 # TRUE
5L == 5 # TRUE
5 + 0i == 5 #TRUE

all(TRUE, FALSE) # FALSE
any(TRUE, FALSE) # TRUE
```

#### Summary Functions
```{r}
x = 1:7
max(x)
min(x)
range(x) # return the min and max
mean(x)
var(x)
sd(x) 
prod(x) # return the product of all elements
sum(x)
```

#### Logical operator
```{r}
!FALSE 
c(FALSE, TRUE) & c(TRUE, TRUE) # elementwise
FALSE && TRUE # 1st element
c(FALSE, TRUE) | c(TRUE, TRUE) # elementwise
FALSE || TRUE # 1st element
xor(FALSE, TRUE) # exclusive OR

x = 5
(x > 5) & (x < 10)
(x > 0) | (x == 0)

value = -3:3
which(value > 0) # return the index of satisfied position
which.max(value) # return the index of maximum
which.min(value) # return the index of minimum

value[which(value > 0)] # get values that are greater than 0
value[which.max(value)] # get the max value
value[value == max(value)]
```

#### Set Operations
```{r}
x = c(1,2,5)
y = c(3,4,5)
union(x,y)
intersect(x,y)
setdiff(x,y)
setequal(x,y)
is.element(1, x) # check if 1 is in set x
1 %in% x # check if 1 is in set x
```

#### General Functions for inspecting a vector
```{r}
x = 1:10
class(x)
length(x)
head(x)
tail(x)
summary(x)
```

Exercise
```{r}
1:1
seq(0, 1, length.out = 10)
rep(c(1,2,3), times = c(1,2,3))

# Four ways to create vector 1,2,3,4,5
c(1,2,3,4,5)
seq(from = 1, to = 5)
1:5
0:4+1

# Generate a random vector of n = 100 elements
a = rnorm(100)
class(a) 
mean(a)
sd(a)
sum(abs(a))
sum(a > 0)
sum(a < 0)
sort(a)[1:3] # get 3 smallest numbers
sort(a, decreasing = TRUE)[1:3] # get 3 largest numbers
```

## Slide 4



## dplyr
filter - select a subset of rows in a data frame  
eg: filter(dat, height > 1, height < 2)  
slice - select rows by position  
eg: slice(dat, 1:10)
arrange - reorder rows  
eg: arrange(dat, year)
    arrange(dat, desc(year))
select - select specific columns
eg: select(dat, year, month, day)
    select(dat, year:day)
    select(dat, -(year:day))
rename - rename some columns
eg: rename(dat, tail_num = tailnum)
distinct - find unique values in a table in that columns
eg: distinct(dat, tailnum)
mutate - add new columns
eg: mutate(dat, gain = year * 2)
transform - mutate allows you to refer to column that you have just created
transmute - keep the newly created variables
eg: transmute(dat, gain = year * 2)

```{r}

```