---
title: "Midterm Project"
author: "Xuanpei Ouyang"
output: pdf_document
subtitle: Stat 133, Fall 2016, Prof. Sanchez
header-includes: \usepackage{float}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set Up: Packages and Functions

You will need to use the following packages: `"stringr"`, `"ggplot2"`, 
`"dplyr"`, and `"readr"`. You can also use any other packages. If you don't 
have any of the packages installed in your computer, install them first 
(outside this Rmd).

```{r}
# use this code chunk to load all the packages that you will be using
# (do not include commands to install the packages)

library(stringr)
library(ggplot2)
library(dplyr)
library(readr)

```

To "import" the functions in your R script file, use the function `source()`

```{r}
# source your functions
source("~/Desktop/STAT 133/stat133-my-repo/midterm_project/stat133-midterm-functions-template.R")
```


# Import Raw Data in R

We are assuming that you already downloaded a copy of the CSV files with the
raw data (you don't need to show the commands you used for this step): 

- `womens-high-jump-raw.csv`
- `mens-high-jump-raw.csv`

Now, use the function `read_csv()` from the package `"readr"` to import the 
data sets in R.

```{r}
# download.file(
#   url = 'https://raw.githubusercontent.com/ucb-stat133/stat133-fall-2016/
#  master/data/womens-high-jump-raw.csv', 
#   destfile = '~/Desktop/STAT 133/stat133-my-repo/midterm_project/
#   womens-high-jump-raw.csv')

# read in womens-high-jump-raw.csv
womens_high_jump_raw = read_csv(
  '~/Desktop/STAT 133/stat133-my-repo/midterm_project/womens-high-jump-raw.csv',
  col_names = TRUE)

# check structure with str()
str(womens_high_jump_raw)

```

```{r}
# download.file(
#   url = 'https://raw.githubusercontent.com/ucb-stat133/stat133-fall-2016/
#   master/data/mens-high-jump-raw.csv', 
#   destfile = '~/Desktop/STAT 133/stat133-my-repo/midterm_project/
#   mens-high-jump-raw.csv')

# read in mens-high-jump-raw.csv
mens_high_jump_raw = read_csv(
  '~/Desktop/STAT 133/stat133-my-repo/midterm_project/mens-high-jump-raw.csv',
  col_names = TRUE)

# check structure with str()
str(mens_high_jump_raw)

```


# Cleaning raw data of women's high jump world records

We'll start by cleaning the women's data set.


### Column `Height`

To clean the contents of column `Height`, use your function to extract height
values to generate a new __numeric__ vector `height` with the values of meters:

```{r}
# vector 'height' with numeric values of height in meters
height = extract_height(womens_high_jump_raw$Height)

```

Clean the column `height` by keeping the meter units and removing the values
in US customary units.

### Column `Athlete`

The column `Athlete` contains the name of the athlete, together with the 
country (inside parenthesis). Use your function to extract the athlete's name
to create a new __character__ vector `athlete`:

```{r}
# vector 'athlete' with character values of athlete's name
athlete = extract_athlete(womens_high_jump_raw$Athlete)

```

Clean the column `athlete` by extracting the name of athlete from the Athlete
column.

Likewise, use your function to extract the country names to create a new __character__ vector `country` that contains the initials of the countries:

```{r}
# character vector 'country' with country abbreviations
country = extract_country(womens_high_jump_raw$Athlete)

```

Extract the column `country` by extracting the country name inside parenthesis
from the Athlete column.

### Column `Date`

The column `Date` contains the dates when the records were established. From
this column you have to create three new vectors: `day`, `month`, and `year`.

- Use your function to extract the day numbers to create a new __numeric__ 
vector `day`:

```{r}
# numeric vector 'day'
day = extract_day(womens_high_jump_raw$Date)

```

Extract the column `day` by extracting the first number (day) from the Date 
column.

- Use your function to extract the month names to create a new __character__ 
vector `month`:

```{r}
# character vector 'month'
month = extract_month(womens_high_jump_raw$Date)

```

Extract the column `month` by extracting the second number (month) from the 
Date column.

- Use your function to extract the year number to create a new __numeric__ 
vector `year`:

```{r}
# numeric vector 'year'
year = extract_year(womens_high_jump_raw$Date)

```

Extract the column `year` by extracting the thrid number (year) from the Date 
column.

Likewise, use your function to reformat date to create a new vector `new_date` 
by reformatting the values in column `Date`:

```{r}
# 'new_date' vector of class "Date"
new_date = reformat_date(womens_high_jump_raw$Date)

```

Reformat the date to a new vector in year-month-day format.

### Column `Place`

Use your function to remove brackets---and the content inside them--- to
create a new character vector `city` from column `Place`:

```{r}
# character vector 'city' with name of city
city = remove_brackets(womens_high_jump_raw$Place)

```

Extract the column `city` by removing the brackets from Place column.

### New vector `gender`

Create a __character__ vector `gender` filled with `"female"` values, having 
length equal to the number of rows in the women's data set.

```{r}
# vector 'gender'
gender = rep("female", nrow(womens_high_jump_raw))

```


### New data frame `womens`

Use the vectors `height`, `athlete`, `gender`, `country`, `city`, `new_date`, 
`day`, `month`, and `year` to build a new data frame `womens`. The column corresponding to `new_date` should have name `"date"`. Do NOT convert
character strings as factors:

```{r}
# data frame 'womens'
womens = data.frame(
  height = height,
  athlete = athlete,
  gender = gender,
  country = country,
  city = city,
  date = new_date,
  day = day,
  month = month,
  year = year,
  stringsAsFactors = FALSE
)

str(womens)
```

Now, I obtain a clean dataset for women's high jump world records with 
9 columns which are height, athlete, gender, country, city, date, day, month
and year. Also I convert all the data column to appropriate data type for
manipulating and ploting data in the later section. 
-----


# Cleaning raw data of men's high jump world records

In this section you will clean the men's data set.


### Column `Height`

To clean the contents of column `Height`, use your function to extract height
values to generate a new __numeric__ vector `height` with the values of meters:

```{r}
# vector 'height' with numeric values of height in meters
height = extract_height(mens_high_jump_raw$Height)
```

Clean the column `height` by keeping the meter units and removing the values
in US customary units.

### Column `Athlete`

The column `Athlete` contains the name of the athlete, together with the 
country (inside parenthesis). Use your function to extract the athlete's name
to create a new __character__ vector `athlete`:

```{r}
# vector 'athlete' with character values of athlete's name
athlete = extract_athlete(mens_high_jump_raw$Athlete)

```

Clean the column `athlete` by extracting the name of athlete from the Athlete
column.

Likewise, use your function to extract the country names to create a new
__character__ vector `country` that contains the initials of the countries:

```{r}
# character vector 'country' with country abbreviations
country = extract_country(mens_high_jump_raw$Athlete)

```

Extract the column `country` by extracting the country name inside parenthesis
from the Athlete column.

### Column `Venue`

The column `Venue` contains the name of the city where the record was 
established. As you can tell from the values in this column, some names contain 
more than just the name of the city (e.g. some include US State, some include
name of country). Use your function to extract the name of the city in 
order to obtain a new __character__ vector `city`:

```{r}
# character vector 'city' with name of city
city = extract_city(mens_high_jump_raw$Venue)

```



### Column `Date`

The column `Date` contains the dates when the records were established. As you 
can tell from the values in this column, they also contain extra characters 
(numbers inside brackets). 

Use your function to remove brackets---and the content inside them--- to
create a new character vector `clean_date`:

```{r}
# vector 'clean_date'
clean_date = remove_brackets(mens_high_jump_raw$Date)

```

Clean the `Date` by removing the brackets followed the data of date

Now take `clean_date` to create three new vectors: `day`, `month`, and `year`.

- Use your function to extract the day numbers to create a new __numeric__ 
vector `day`:

```{r}
# numeric vector 'day'
day = extract_day(clean_date)

```

Extract the column `day` by extracting the first number (day) from the Date 
column.

- Use your function to extract the month names to create a new __character__ 
vector `month`:

```{r}
# character vector 'month'
month = extract_month(clean_date)

```

Extract the column `month` by extracting the second number (month) from the 
Date column.

- Use your function to extract the year number to create a new __numeric__ 
vector `year`:

```{r}
# numeric vector 'year'
year = extract_year(clean_date)

```

Extract the column `year` by extracting the third number (year) from the 
Date column.

Likewise, use your function to reformat date to create a new vector `new_date` 
by reformatting the values in column `Date`:

```{r}
# 'new_date' vector of class "Date"
new_date = reformat_date(clean_date)

```

Reformat the date to a new vector in year-month-day format.

### New vector `gender`

Create a __character__ vector `gender` filled with `"male"` values, having 
length equal to the number of rows in the men's data set.

```{r}
# vector 'gender'
gender = rep("male", nrow(mens_high_jump_raw))

```


### New data frame `mens`

Use the vectors `height`, `athlete`, `gender`, `country`, `city`, `new_date`, 
`day`, `month`, and `year` to build a new data frame `mens`. The column 
corresponding to `new_date` should have name `"date"`. Do NOT convert
character strings as factors:

```{r}
# data frame 'mens'
mens = data.frame(
  height = height,
  athlete = athlete,
  gender = gender,
  country = country,
  city = city,
  date = new_date,
  day = day,
  month = month,
  year = year,
  stringsAsFactors = FALSE
)

str(mens)
```

Here, I obtain a clean dataset for men's high jump world records with 
9 columns which are height, athlete, gender, country, city, date, day, month
and year. Also, I convert all the data column to appropriate data type for
manipulating and ploting data in the later section. 


-----

# Data Manipulation

Create a new data frame `records` by merging (or "stacking") the data frames 
`womens` and `mens`. Only the columns `gender` and `country` should be 
converted as R factors. Make sure that the column `date` is of class `"Date"`.

```{r}
# new data frame 'records'
records = rbind(womens, mens)
records$gender = as.factor(records$gender)
records$country = as.factor(records$country)

str(records)
```

I merge the two data frame for women and men to create a new data frame called
records, which contains the same columns as men's and women's dataset. Here, I
convert the `gender` and `country` to R factors for plotting purpose.

Use functions in `"dplyr"` to compute the following frequency tables. Assign
each table to its own object.

### Frequency Table for number of records per country

```{r}
# number of records per country
# object: per_country
per_country = records %>%
  group_by(country) %>%
  summarise(records = n())

# invoke head() on per_country
head(per_country)
```

`per_country` is a frequency table containing the numbers of high jump records 
per country from year 1912 to 1993 sorted by the alphabetic order of country
name.

```{r}
# number of records per country in descending order
# object: per_country_desc
per_country_desc = per_country %>%
  arrange(desc(records))

# invoke head() on per_country_desc
head(per_country_desc)
```

`per_country_desc` is a frequency table containing the numbers of high jump 
records per country from year 1912 to 1993 sorted by the descending order of 
records counts. And we can see that USA has the most high jump records 
during 1912 and 1993.

```{r}
# number of records per year in descending order
# object: per_year_desc
per_year_desc = records %>%
  group_by(year) %>%
  summarise(records = n()) %>%
  arrange(desc(records))
  
# invoke head() on per_year_desc
head(per_year_desc)
```

`per_year_desc` is a frequency table containing the numbers of high jump records 
per year from year 1912 to 1993 sorted by the descending order of records 
counts. And we can see that the highest number of high jump records were made
in 1961.

```{r}
# number of records by gender
# object: by_gender
by_gender = records %>%
  group_by(gender) %>%
  summarise(records = n())

# invoke head() on by_gender
head(by_gender)
```

`by_gender` is a frequency table containing the numbers of high jump records 
seperated by gender from year 1912 to 1993 and sorted by the descending order 
of records counts.

Here is one more challenge that involves some data wrangling with `year`.
The objective is to compute the number of records per decade (regardless of
gender):

```{r}
# number of records per decade (1920-1929, 1930-1939, 1940-1949, etc)
# object: per_decade
per_year = records %>%
  group_by(year) %>%
  summarise(records = n()) %>%
  arrange(year)

intervals = seq(from = 1910, to = 2000, by = 10)
decade_records = integer(9)

# Use for loop to select appropriate years and count the sum of number 
# of records in those selected years
for (i in 1:length(decade_records)){
  filtered_year = filter(per_year, year >= intervals[i] , year < intervals[i+1])
  filtered_records = filtered_year[,2]
  decade_records[i] = sum(filtered_records)
}
 
decade_names = c("1910-1919", "1920-1929", "1930-1939","1940-1949","1950-1959",
            "1960-1969","1970-1979","1980-1989","1990-1999")

# create a data frame with appropriate names per_decade
per_decade = data.frame(
  decade = decade_names,
  records = decade_records,
  stringsAsFactors = FALSE
)
# invoke head() on per_decade
head(per_decade)
```

`per_decade` is a frequency table containing the number of records per decade 
regardless of gender and country and is sorted by earliest to latest year.
The year interval used here is 1910-1929, 1920-1929, 1930-1939, 1940-1949, 
1950-1959, 1960-1969, 1970-1979, 1980-1989, 1990-1999.
And, we can see that the number of high jump world records made during 
1980-1989 is the highest.

-----

# Data Visualization

Use the `records` data frame, and functions in `"ggplot2"`, to create charts 
similar to those displayed in the PDF with the instructions for this project:

- Include a title on all graphics
- Include a description for each graphic

```{r}
# step-line plot, with points, showing the progression of records
# (lines colored by gender)
# (the default ggplot color values are: '#00BFC4', '#F8766D')
ggplot(records, aes(x = date, y = height, group = gender)) +
  geom_point(aes(color = gender), show.legend = FALSE) + 
  geom_step(aes(color = gender), show.legend = FALSE) + 
  annotate("text", x = as.Date("1990-06-01"), y = 1.90, label = "women", 
           color = "#F8766D") + 
  annotate("text", x = as.Date("1990-06-01"), y = 2.25, label = "men", 
           color = "#00BFC4") + 
  ggtitle("High Jump World Records Progression (by gender)") +
  theme_minimal() 
```

### High Jump World Records Progression (by gender)
The plot above is a step-line graph, with points, of the progression of the 
high jump world records by gender. This graph shows the relationship between 
world records and the time stamp of the creation of world records. Based on the 
dataset of high jump world records for men and women, I plots each data point,
joints them up with a step line and also color them seperately. From the plot,
we can see that the overall pattern between the world records value and date 
has a postivie association. Also, these two variables has approximately linear
association. Because of the special properties of world records that it can 
only increase, the relationship between world record and the time is mostly 
likely to be positive.
For both men's and women's high jump world records, we can see that world 
records for both men and women plateaued several times, especially from 1941 
to 1952 and from 1961 to 1971, then increase rapidly after the plateau 
possibly due to some breakthroughs in high jump technique. 
Also, the progression step-line for men's world records is absolutely 
above the progression step-line for women's world records indicates that men
may have advantage in high jump because of some physical reasons.

```{r}
# step-line plot, with points, showing the progression of records
# (facetting by gender)
ggplot(records, aes(x = date, y = height)) +
  geom_point(aes(color = gender), show.legend = FALSE) + 
  geom_step(aes(color = gender), show.legend = FALSE) + 
  ggtitle("High Jump World Records Progression (by gender)") + 
  facet_grid(~ gender)
```

### High Jump World Records Progression (by gender)
The plot is a step-line graph, with points, for showing the progression of 
high jump world records by gender. Here, I sliced the dataset by gender and 
plot the relationship between world records and date by gender in sub-panels
in the same plot. From the plot, we can compare the progression of men's and 
women's high jump world records over time easily. The overall trends of their 
progressions are both postive but men's world records is always higher than 
women's world records. Also, by the end of 1977, women's world record reach 
men's first world records created around 1912.

```{r}
# step-line plot, with points, showing the progression of women records
# (points colored by country)
womens = filter(records, gender == "female")
Country = womens$country
ggplot(womens, aes(x = date, y = height)) +
  geom_step(color = "gray", size = 1) +
  geom_point(aes(color = Country)) + 
  ggtitle("Women's High Jump World Records Progression") + 
  theme_minimal()
```

### Women's High Jump World Records Progression
This plot is a step-line graph, with points, showing the progression of women 
records. Also, in this plot, I made data points colored by country to 
visualize the corresponding country for each record. From the plot, we can see 
that world records are spread among different countries. And in some small 
time interval, athelets from the same country created world records 
consecutively, such as Romanian from 1958 to 1962, Great Britain from 1975 to
1980.

```{r}
# step-line plot, with points, showing the progression of men records
# (points colored by country)
mens = filter(records, gender == "male")
Country = mens$country
ggplot(mens, aes(x = date, y = height)) +
  geom_step(color = "gray", size = 1) +
  geom_point(aes(color = Country)) + 
  ggtitle("Men's High Jump World Records Progression") + 
  theme_minimal()
```

### Men's High Jump World Records Progression
This plot is a step-line graph, with points, showing the progression of men 
records. Also, I made data points colored by country to visualize the 
corresponding country for each record. We can see that before year 1960, most 
of the world records were created by USA athlets and during year 1960 to 1967, 
most of the world records were created by Russia athlets. And overall, before
1980, men's world records only came from USA and Russia and after that, other 
countries such as China, Cuba, West Germany, East Germany, Poland, Soviet Union
and Sweden. 


-----

# Model Fitting

Now let's do some basic model fitting with a simple regression analysis.


## Women's regression model

Subset the data `records` for those observations of gender `female`, and 
use the function `lm()` to compute a linear model by regressing `height` on
`year`:

```{r}
# Model for women
# regression of 'height' on 'year'
women = subset(records, gender == "female")
height = women$height
year = women$year
height_on_year_female = lm(height ~ year, data = women)
height_on_year_female
plot(height_on_year_female)
```

Use the object `"lm"` object in the previous step to compute a "prediction"
of what the women's high jump world record could have been in years: 
2000, 2004, 2008, 2012, and 2016.

```{r}
# predictions for olympic years
predict_year = c(2000, 2004, 2008, 2012, 2016)
years_for_prediction = data.frame(
  year = predict_year)
predict_height_women = predict.lm(height_on_year_female, years_for_prediction)
prediction_women = data.frame(
  year = predict_year,
  predict_height = predict_height_women
)
prediction_women
ggplot(womens, aes(x = year, y = height)) + 
  geom_point() +
  ggtitle("Women's High Jump World Record Linear Regression Line") + 
  stat_smooth(method = "lm", col = "red")
```

I compute a linear model for women's high jump world record by regressing
`height` on `year`. Then I use the computed linear model to predict the 
women's high jump world record and the prediction output is displayed in 
the above table. Also, to visualize how well the regression line fits, I plot
the regression line on scatter plot. We can see that the regression
line fits well with the original dataset. 


## Men's regression model

Subset the data `records` for those observations of gender `male`, and 
use the function `lm()` to compute a linear model by regressing `height` on
`year`:

```{r}
# Model for men
# regression of 'height' on 'year'
men = subset(records, gender == "male")
height = men$height
year = men$year
height_on_year_male = lm(height ~ year, data = men)
height_on_year_male
plot(height_on_year_male)
```

Use the object `"lm"` object in the previous step to compute a "prediction"
of what the men's high jump world record could have been in years: 
2000, 2004, 2008, 2012, and 2016.

```{r}
# predictions for olympic years
years_for_prediction = data.frame(
  year = predict_year)
predict_height_men = predict.lm(height_on_year_male, years_for_prediction)
prediction_men = data.frame(
  year = predict_year,
  predict_height = predict_height_men
)
prediction_men
ggplot(mens, aes(x = year, y = height)) + 
  geom_point() +
  ggtitle("Men's High Jump World Record Linear Regression Line") + 
  stat_smooth(method = "lm", col = "red")
```

I compute a linear model for men's high jump world record by regressing
`height` on `year`. Then I use the computed linear model to predict the 
men's high jump world record and the prediction output is displayed in 
the above table. Also, to visualize how well the regression line fits, I plot
the regression line on scatter plot. We can see that the regression
line fits well with the original dataset. 


-----

## Extra Credit

This part of the project is optional. If you provide a satisfactory solution,
you will get extra credit.

Consider the first women's world record which is 1.46 meters. This value
corresponds to "4 ft 9 1/2 in". Likewise, a height value of 2.04 meters can be 
expressed in US customary units as "6 ft 8 3/8 in".

The goal is to write a function that takes a height value (in meters), and 
which returns a character vector in US customary units. In other words, 
if you pass the value `1.46` to your function, the output will be the string 
`"4 ft 9 1/2 in"`.

You should be able to apply your function on the entire vector `height` to get 
the corresponding strings in US customary units. Your results should match the 
values in wikipedia's html tables.

Don't write a long function. You can use as many auxiliary short functions as
necessary. All the functions must be well documented with a general description
for the function's purpose, the arguments, and the output. Include these 
functions in the same `.R` script file with the other functions (those 
used to clean the columns of the raw data sets).

Your code should be accompanied with descriptions of how you approached this
problem, and the decisions you made to implement it in that way.
```{r}
# Code for testing
height_in_meters = records$height
converted_height_string = character(length(height_in_meters))

for (i in 1:length(height_in_meters)){
  converted_height_string[i] = 
    convert_height_units_string(height_in_meters[i])
}

converted_height_string
```
