---
title: "HW02 - Data tables and graphics"
subtitle: "Stat 133, Fall 2016, Prof. Sanchez"
author: "Xuanpei Ouyang"
output: html_document
---

This assignment has 4 main purposes:

- practicing reading data tables
- working with data frames, vectors and factors
- practice basic manipulation of data structures
- creation of simple graphics with `"ggplot2"`

Download this file or open an `Rmd` file in RStudio and copy the
contents of this file.


## Data set Camping Tents

Start by practicing reading tables in R. The data for this assignment is in the 
file `camping-tents.csv` (available in the folder `data/` in the 
github repository of the course)

Use `download.file()` to download the file `camping-tents.csv` in your computer.
You'll have to specify your own destination (`destfile`)
e.g. in my case I'm specifying: `"~/Downloads/camping-tents.csv"`
(your `destfile` will probably be different!)

```{r eval = FALSE}
download.file(
  url = 'https://raw.githubusercontent.com/ucb-stat133/stat133-fall-2016/master/data/camping-tents.csv', 
  destfile = '~/camping-tents.csv')
```

- __Importing A:__ Use `read.csv()` to import `camping-tents.csv` and assign 
it to an object called `tdf1`

```{r}
# import using 'read.csv()'
tdf1 = read.csv('~/camping-tents.csv')
```


- __Importing B:__ Use `read.csv()` to import `camping-tents.csv` and assign 
it to an object called `tdf2`. In this case, strings must NOT be converted into 
factors!
```{r}

# import using 'read.csv()' without converting strings into factors
tdf2 = read.csv('~/camping-tents.csv', stringsAsFactors = FALSE)

```


- __Importing Option 3:__ Create a character vector `col_class` with the 
following classes for the columns (to be used with argument `colClasses`)
  - name:     `character`
  - brand:    `character`
  - price:    `numeric`
  - weight:   `numeric`
  - height:   `integer`
  - bestuse:  `factor`
  - seasons:  `factor`
  - capacity: `factor`

Use `read.table()` to import `camping-tents.csv` and assign it to 
an object called `tdf3`. Pass the vector `col_class` to the argument 
`colClasses`, and specify the apropriate field separator, and header.

```{r}
# import using 'read.csv()' specifying `colClasses

col_class = c(
  'character', 
  'character', 
  'numeric', 
  'numeric', 
  'integer', 
  'factor', 
  'factor', 
  'factor'
)

tdf3 = read.csv('~/camping-tents.csv', header = TRUE, sep = ',', colClasses = col_class)

```


- __Importing Option 4:__ Use `read_csv()` from package `"readr"` to import 
`camping-tents.csv` and assign it to an object called `tents`.
Use the argument `col_types` to specify column types:
  - name:     `character`
  - brand:    `character`
  - price:    `numeric`
  - weight:   `numeric`
  - height:   `integer`
  - bestuse:  `factor`
  - seasons:  `factor`
  - capacity: `factor`


```{r}
# import using 'read_csv()'
library('readr')

col_types_list = list(
  name = col_character(),
  brand = col_character(),
  price = col_double(),
  weight = col_double(),
  height = col_integer(),
  bestuse = col_factor( c('Backpacking', 'Mountaineering', 'Carcamping') ),
  seasons = col_factor( c('3-season', '4-season', '3-4-season') ),
  capacity = col_factor( c('1-person', '2-person', '3-person', '4-person', '6-person') )
)

tents = read_csv('~/camping-tents.csv', col_types = col_types_list)
```


-----

## Inspecting data frames

Once you have your data frame `tents`, the next step 
involves performing a first exploration.

```{r}
# inspect the data structure of 'tents'
str(tents)
summary(tents)

# display the names of columns in 'tents'
colnames(tents)

# take a look at the first 5 rows
head(tents, 5)

# take a look at the last 3 rows
tail(tents, 3)
```


### Quantitative Variables:

Exploring the quantitative variables. Instead of using the `summary()` function, 
find functions that allow you to get the following summaries for `price`

- minimum value
- first quartile (i.e. 25th percentile)
- third quartile (i.e. 75th percentile)
- 90th percentile
- median value (i.e. 50th percentile)
- mean value
- maximum value
- inter-quartile range (IQR)
- standard deviation

```{r}
# your code

# find minimum value
min(tents$price)
# find first, thrid quartile, 90th percentile
quantile(tents$price, c(0.25, 0.75, 0.9))
# find median value (two ways)
median(tents$price)
quantile(tents$price, c(0.5))
# find mean value
mean(tents$price)
# find maximum value
max(tents$price)
# inter-quartile range
IQR(tents$price)
# standard deviation
sd(tents$price)
```


Add a new variable `weight_lbs` to the data frame `tents` for weight in pounds.
Keep in mind that `weight` is given in grams.
```{r}
# add weight_lbs
grams_to_lbs = 0.00220462262185 * tents$weight
tents$weight_lbs = grams_to_lbs
```

Add a new variable `height_in` to the data frame `tents` for height in inches.
Keep in mind that `height` is given in centimeters.
```{r}
# add height_in
centimeters_to_inches = 0.39370079 * tents$height
tents$height_in = centimeters_to_inches
```

Write R code (using bracket notation) to answer the following questions:
```{r}
# how many tents have a price less than or equal to $300
length(tents$price[tents$price <= 300])

# how many tents have a price between $300 and $400
# (including both $300 and $400 prices)
length(tents$price[(tents$price >= 300) & (tents$price <= 400)])

# what's the name of the tent with maximum price
tents$name[tents$price == max(tents$price)]

# select the data of tents with 
# price > $400 AND weight < 1500 grams
pw_row_logical_index = (tents$price > 400) & (tents$weight < 1500)
tents[pw_row_logical_index,]

# calculate the 90th percentile for height and 
# assign it to the object 'height_p90'
height_p90 = quantile(tents$height, c(0.9))

# calculate the 90th percentile for weight and 
# assign it to the object 'weight_p90'
weight_p90 = quantile(tents$weight, c(0.9))

# select the data of tents with 
# height > height_p90 AND weight > weight_p90
p90_row_logical_index = (tents$height > height_p90) & (tents$weight > weight_p90)

```


-----

## More Manipulations

Use `subset()` to obtain a new data frame called `rei` containing the data of 
those tents with brand `'rei'`; and inspect its dimension:
```{r}
# rei data frame
rei = subset(tents, brand == 'rei')
dim(rei)
```


Use `cut()` to create a factor `weight_cut` by using the
breaking points and labels according to the following table:
```
-----------------------
     intervals | labels
---------------|-------
     (0, 1000] |  1kg
  (1000, 2000] |  2kg
  (2000, 3000] |  3kg
  (3000, 4000] |  4kg
  (4000, 5000] |  5kg
  (5000, 6000] |  6kg
  (6000, 7000] |  7kg
  (7000, 8000] |  8kg
  (8000, 9000] |  9kg
----------------------
```

```{r}
# your weight_cut factor
intervals = seq(from = 0, to = 9000, by = 1000)
my_label = c('1kg','2kg','3kg','4kg','5kg','6kg','7kg','8kg','9kg')
weight_cut = cut(tents$weight, breaks = intervals, labels = my_label)
```

Verify that the frequencies given by `table(weight_cut)` are:
```
 1kg 2kg 3kg 4kg 5kg 6kg 7kg 8kg 9kg 
   3  24  30   7   7   2   9   3   3 
```

```{r}
# your verification
table(weight_cut)
```


-----

## Statistical graphics of Quantitative Variables

Use `"ggplot2"` to obtain the following plots for variable `price`: 
(look at each graphic carefully and see what types 
of distribution patterns show each variable)

- histogram
- boxplot
- density curve

```{r}
# your plots
library(ggplot2)
price = tents$price

# histogram
ggplot(tents, aes(x = price)) + geom_histogram()

# boxplot group ? by brand
ggplot(tents, aes(factor(brand), price)) + geom_boxplot()

# density curve
ggplot(tents, aes(x = price)) + geom_density()

```


### Plot challenge

Here's a small challenge: Obtain a scatter plot of `weight` and `height` 
such that:

- label of x-axis is: `Weight (in grams)`
- label of y-axis is: `Height (in csm)`
- use `bestuse` as a factor for the color argument
- Add alpha transparency `alpha = 0.7` to the color of the points
- The symbol of points are filled triangles
- x-axis ranges from 0 to 10000
- y-axis ranges from 80 to 220
- Include a title

```{r}
# your scatter plot
ggplot(tents, aes(x = weight, y = height)) + 
  geom_point(aes(colour = factor(bestuse)), alpha = 0.7, shape=17) +
  labs( title = 'Scatter Plot for Weight and Height', x = 'Weight (in grams)', y ='Height (in csm)') +
  coord_cartesian(xlim=c(0,10000), ylim=c(80,220))
```


-----

## Statistical graphics of Qualitative Variables

Now let's focus on the qualitative variables:

1. `brand`
2. `bestuse`
3. `seasons`
4. `capacity`

```{r}

  
# what is the brand with less number of tents
ggplot(tents, aes(x=brand)) + 
  geom_bar() 
# Based on the graph, sierra-designs has the less number of tents

# are there any tents of brand 'rei'?
# and if so, how many?
any(tents$brand == 'rei')
length(tents$brand[tents$brand == 'rei'])
# Yes, 22

# are there any tents of brand 'mountain-hardwear'
# and if so, how many?
any(tents$brand == 'mountain-hardwear')
# No

# how many 'rei' tents are intended to be
# used ('bestuse') for Mountaineering?
ggplot(tents, aes(x = brand, fill = bestuse)) + 
  geom_bar() 
length(tents$name[(tents$brand == 'rei') & (tents$bestuse == 'Mountaineering')])
# 2 rei tents are intended to be used for 'Mountaineering'

# what unique brands have tents intended to be used for 'Mountaineering'
# Based on the graph, marmot, msr, north-face and rei have tents intended to be 
# used for 'Mountaineering'

```


### Plot challenge: 

Make a barplot for 'brand' such that:
- bars are horizontally oriented
- bars are arranged in decreasing order
- brand labels are perpendicular to the y-axis
  (i.e. labels horizontally oriented)
- modify margins in order to have enough room for labels 
- x-axis ranges from 0 to 25
- include a title

```{r}
# your bar-chart

#brand_copy = unique(tents$brand)

#count = c(le[tents$brand == brand_copy[1]]),
#          length(tents$name[tents$brand == brand_copy[2]]),
#          length(tents$name[tents$brand == brand_copy[3]]),
#          length(tents$name[tents$brand == brand_copy[4]]),
#          length(tents$name[tents$brand == brand_copy[5]]),
#          length(tents$name[tents$brand == brand_copy[6]]),
#          length(tents$name[tents$brand == brand_copy[7]]),
#          length(tents$name[tents$brand == brand_copy[8]])
#)

#tents$unique_
ggplot(tents, aes(x = brand)) + 
  geom_bar() + 
  coord_flip() + 
  scale_y_continuous( limits = c(0, 25) )


  #
  #labs( title = "Counts for Different Brand Tents" ) +
  #theme(
  #  text = element_text(margin = margin(5))
  #)
  # arrange bar in  a decreasing order
```


### Bivariate Plots

Use `geom_boxplot()` to plot the distribution of `weight` conditional to:
(get a boxplot for each categorical variable)

- brand
- bestuse
- capacity

```{r}
# your boxplots
ggplot(tents, aes(factor(brand), price)) + geom_boxplot()
ggplot(tents, aes(factor(bestuse), price)) + geom_boxplot()
ggplot(tents, aes(factor(capacity), price)) + geom_boxplot()
```


Make a scatter-plot of `height` and `weight`, using 
`capacity` as a factor for the color argument
```{r}
# scatter-plot of `height` and `weight` (color `capacity`)
ggplot(tents, aes(x = weight, y = height, color = capacity)) + geom_point()

```


