---
title: "Lab 3: Intro to the package ggplot2"
subtitle: "Stat 133, Fall 2016"
author: "Gaston Sanchez"
header-includes: \usepackage{float}
output: html_document
---

> ### Learning Objectives
>
> - Introduction to the R package "ggplot2"
> - Understanding the grammar of ggplot
> - Plots for one variable
> - Scatter plots
> - Time line
> - Faceting

------


## Graphics in R

R has an amazing ecosystem of graphing capabilities to produce static plots.
As for dynamic/interactive plots, R is very limited, although there have 
been various packages developed in recent years to provide some interesting solutions.

In this session, we will focus on the most popular R package to create 
graphics: `"ggplot2"`
 

## Introduction to ggplot2 (by Karthik Ram)

Walk through Karthik's slides
[https://github.com/karthik/ggplot-lecture](https://github.com/karthik/ggplot-lecture)

Slides: https://github.com/karthik/ggplot-lecture/blob/master/ggplot.pdf

``` {r}
# install.packages('ggplot2', dependencies = TRUE)
library(ggplot2)
```


IRIS

```{r}
head(iris)
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width))
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) + geom_point()

iris_scatter = ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width))
iris_scatter + geom_point()

qplot(iris$Sepal.Length, iris$Sepal.Width)

iris_scatter + geom_point(size = 10)
iris_scatter + geom_point(size = 0.1)

iris$Species
iris_scatter = ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species))
iris_scatter + geom_point()

iris_scatter = ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width))
iris_scatter + geom_point(aes(shape = Species, color = Species)) # color in geom_point
```


DIAMONDS

```{r}
head(diamonds)
diamonds_sub <- diamonds[sample(1:dim(diamonds)[1], 1000),]
diamond_scatter <- ggplot(data = diamonds_sub, aes(carat, price, color = color))
diamond_scatter + geom_point()

```


IRIS2
```{r}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) + 
  geom_point() +
  facet_grid(Species ~ .)

ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) + 
  geom_point() +
  facet_grid(. ~ Species)

```

```{r}
library(RColorBrewer)
display.brewer.all()

```


IRIS3
```{r}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) + 
  geom_point() +
  facet_grid(. ~ Species) +
  scale_color_manual(values = c('blue', 'purple', 'green'))

ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) + 
  geom_point() +
  facet_grid(. ~ Species) +
  scale_color_manual(values = c('blue', 'purple', 'green')) +
  geom_smooth(method = 'lm')

ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) + 
  geom_point(aes(shape = Species)) +
  scale_color_manual(values = c('blue', 'purple', 'green')) +
  geom_smooth(method = 'loess')

```

DIAMOND2
ggplot with carat vs. depth, categorized by color, points scaled to 0.5, fit 'loess' (local polynomical regression), facetted by columns)
```{r}
# YOUR CODE
```





SOLUTION:
```{r}
ggplot(data = diamonds_sub, aes(carat, price, color = color)) + 
  geom_point(size = 0.5) + 
  geom_smooth(method = 'loess') + 
  facet_grid(. ~ color)
```

# Legend key/position, axes titles

# ?theme

# install.packages('ggthemes')
# library(ggthemes)

-----

  
  
  
### References

- [RStudio's ggplot2 cheat sheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)
- [ggplot2 official documentation](http://docs.ggplot2.org/current/)
- [Using R package plotly's with ggplot2 to create interactive, online ggplot2 charts](https://plot.ly/ggplot2/)
- [ggplot2: Elegant Graphics for Data Analysis](http://ggplot2.org/book/) (by Hadley Wickham)
- [R Graphics Cookbook](http://shop.oreilly.com/product/0636920023135.do) (by Winston Chang)
- [stack overflow questions tagged with ggplot2](http://stackoverflow.com/questions/tagged/ggplot2)

-----
